---
title: "Alpha diversity analysis"
author: "Elisa Rubio"
date: "2022-03-03"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(cache=FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE, results ='markup')
options(warn=-1)
```

```{r message=FALSE}
library(tidyverse)
library(kableExtra)
library(vegan)
library(ggpubr)
```

## Load summarized and rarefied data

```{r}
load("output/summarized_ariba.RDA")
load("output/summarized_groot.RDA")
load("output/rarefied_ariba.RDA")
load("output/rarefied2500_groot.RDA")
metadata<-read_csv("data/Metadata/metadata.csv")
```

## Create functions:

### Calculate apha diversity indexes:

```{r}
alpha_div<-function(d_rpkm, d_counts){
shannon<-diversity(d_rpkm[ ,-1], MARGIN = 2, index="shannon")
simpson<-diversity(d_rpkm[, -1], MARGIN = 2, index="simpson")
invsimpson<-diversity(d_rpkm[, -1], MARGIN = 2, index="invsimpson")
nARG<-specnumber(d_rpkm[, -1], MARGIN = 2)
fisher<-fisher.alpha(d_counts[ ,-1], MARGIN = 2) ##we obtain fisher index from counts (not possible to calculate for rpkm values)
metadata<-read_csv("data/Metadata/metadata.csv")%>%
  filter(SampleID %in% names(d_rpkm))%>%
  mutate(Shannon=shannon, Simpson=simpson, InvSimpson= invsimpson, nARG=nARG, Fisher=fisher)}
```

### Plot a diversity index for a clinical variable

```{r}
alpha_div_plot<-function(data, clin_var, alpha_var){
  clin_var<-enquo(clin_var)
  alpha_var<-enquo(alpha_var)
ggplot(data, aes(x=!!clin_var, y=!!alpha_var))+
  geom_boxplot(aes(colour=!!clin_var))+
  geom_jitter(width = 0.2, aes(colour=!!clin_var))+
  stat_compare_means()+
  theme_classic()+
  theme(legend.position = "none")}
```

### Plot all diversity indexes for a clinical variable

```{r}
alpha_div_plot_all<-function(data, clin_var){
clin_var<-enquo(clin_var)
data%>%
  pivot_longer(cols=c(Shannon, Simpson, InvSimpson, nARG, Fisher), names_to = "Alpha_Index", values_to = "Alpha_value")%>%
  ggplot(aes(x=!!clin_var, y=Alpha_value))+
  geom_boxplot(aes(colour=!!clin_var))+
  geom_jitter(width = 0.2, aes(colour=!!clin_var))+
  facet_wrap(vars(Alpha_Index), ncol = 3, scales = "free_y")+
  stat_compare_means(label="p.format", label.x = 1, label.y.npc = 0.9)+
  theme_bw()+
  labs(y="Alpha diversity", caption = "P values calculated by Wilcoxon test", title = clin_var)+
  theme(legend.position = "none", axis.title.x=element_blank(),
        axis.text.x = element_text(angle=45, hjust = 1))}

```

### Plot correlation between diversity indexes and quantitative clinical variables

```{r}
alpha_div_cor_all<-function(data, clin_var){
clin_var<-enquo(clin_var)
data%>%
  pivot_longer(cols=c(Shannon, Simpson, InvSimpson, nARG, Fisher), names_to = "Alpha_Index", values_to = "Alpha_value")%>%
  ggplot(aes(x=!!clin_var, y=Alpha_value))+
  geom_point(color='#2980B9', size = 2) + 
  geom_smooth(method=lm, color="black")+
  facet_wrap(vars(Alpha_Index), ncol = 3, scales = "free_y")+
  stat_cor(method="spearman", label.x.npc = "left", label.y.npc = 0.9)+
  theme_bw()+
  labs(y="Alpha diversity", caption = "Spearman correlation", title = clin_var)+
  theme(legend.position = "none", axis.title.x=element_blank(),
        axis.text.x = element_text(angle=45, hjust = 1))}
```


## ARIBA

```{r ariba_alpha_rpkmdiv}
ariba_alpha_rpkm<-alpha_div(ariba_rpkm, ariba_counts)
ariba_alpha_rar<-alpha_div(ariba_rar, ariba_rar)
```


### HIV status

```{r}
alpha_div_plot_all(ariba_alpha_rpkm, HIV_Status)+labs(subtitle = "Ariba RPKM")
alpha_div_plot_all(ariba_alpha_rar, HIV_Status)+labs(subtitle = "Ariba Rarefied counts")
```

### Profile

```{r}
alpha_div_plot_all(ariba_alpha_rpkm, Profile)+labs(subtitle = "Ariba RPKM")
alpha_div_plot_all(ariba_alpha_rar, Profile)+labs(subtitle = "Ariba Rarefied counts")
```


### Risk Group 2

```{r}
alpha_div_plot_all(ariba_alpha_rpkm, RiskGroup2)+labs(subtitle = "Ariba RPKM")
alpha_div_plot_all(ariba_alpha_rar, RiskGroup2)+labs(subtitle = "Ariba Rarefied counts")
```


### CD4 (absolute)

```{r}
alpha_div_cor_all(ariba_alpha_rpkm, CD4)+labs(subtitle = "Ariba RPKM")
alpha_div_cor_all(ariba_alpha_rar, CD4)+labs(subtitle = "Ariba Rarefied counts")
```


### CD4 (relative)

```{r}
alpha_div_cor_all(ariba_alpha_rpkm, CD4_relative)+labs(subtitle = "Ariba RPKM")
alpha_div_cor_all(ariba_alpha_rar, CD4_relative)+labs(subtitle = "Ariba Rarefied counts")
```

### Nadir CD4

```{r}
alpha_div_cor_all(ariba_alpha_rpkm, Nadir_CD4)+labs(subtitle = "Ariba RPKM")
alpha_div_cor_all(ariba_alpha_rar, Nadir_CD4)+labs(subtitle = "Ariba Rarefied counts")
```


## Groot

```{r groot_alpha_rpkmdiv}
groot_alpha_rpkm<-alpha_div(groot_rpkm, groot_counts)
groot_alpha_rar<-alpha_div(groot_rar, groot_rar)
```


### HIV status

```{r}
alpha_div_plot_all(groot_alpha_rpkm, HIV_Status)+labs(subtitle = "groot RPKM")
alpha_div_plot_all(groot_alpha_rar, HIV_Status)+labs(subtitle = "groot Rarefied counts")
```

### Profile

```{r}
alpha_div_plot_all(groot_alpha_rpkm, Profile)+labs(subtitle = "Groot RPKM")
alpha_div_plot_all(groot_alpha_rar, Profile)+labs(subtitle = "Groot Rarefied counts")
```


### Risk Group 2

```{r}
alpha_div_plot_all(groot_alpha_rpkm, RiskGroup2)+labs(subtitle = "Groot RPKM")
alpha_div_plot_all(groot_alpha_rar, RiskGroup2)+labs(subtitle = "Groot Rarefied counts")
```


### CD4 (absolute)

```{r}
alpha_div_cor_all(groot_alpha_rpkm, CD4)+labs(subtitle = "Groot RPKM")
alpha_div_cor_all(groot_alpha_rar, CD4)+labs(subtitle = "Groot Rarefied counts")
```


### CD4 (relative)

```{r}
alpha_div_cor_all(groot_alpha_rpkm, CD4_relative)+labs(subtitle = "Groot RPKM")
alpha_div_cor_all(groot_alpha_rar, CD4_relative)+labs(subtitle = "Groot Rarefied counts")
```

### Nadir CD4

```{r}
alpha_div_cor_all(groot_alpha_rpkm, Nadir_CD4)+labs(subtitle = "Groot RPKM")
alpha_div_cor_all(groot_alpha_rar, Nadir_CD4)+labs(subtitle = "Groot Rarefied counts")
```

