---
title: "09_group matrixes"
author: "Elisa Rubio"
date: "2022-11-09"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(cache=FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE, results ='markup')
options(warn=-1)
```

```{r message=FALSE}
library(tidyverse)
library(kableExtra)
library(vegan)
library(ggpubr)
library(glue)
library(patchwork)
library(ggstatsplot)
library(broom)
```


#### Load data

```{r}
load("output/summarized_ariba.RDA")
load("output/rarefied_ariba.RDA")
metadata<-read_csv("data/Metadata/metadata.csv")
```


## Generate matrix counts grouped by a level Gene family

```{r}
ariba_rpkm_genefamily<-ariba_rpkm%>%left_join(refname_all_ariba, "ref_name")%>%
  select(`AMR Gene Family`, starts_with("Sample"))%>%
  group_by(`AMR Gene Family`)%>%
  summarise(across(everything(), sum),
            .groups = 'drop')

ariba_rar_genefamily<-ariba_rar%>%left_join(refname_all_ariba, "ref_name")%>%
  select(`AMR Gene Family`, starts_with("Sample"))%>%
  group_by(`AMR Gene Family`)%>%
  summarise(across(everything(), sum),
            .groups = 'drop')

```

## Generate matrix counts grouped by antibiotic and resistance mechanism

Problem: one AMR may confer resistance to more than one antibiotic

Option1. We count each ATB separately (Reads will be counted more than once if they confer resistance to more than one antibiotic)
We group all beta-lactamic antibiotics in one group andthe penam/penem group in another

```{r}
ariba_rpkm_drugclass<-ariba_rpkm%>%left_join(refname_all_ariba, "ref_name")%>%
  select(`Drug Class`, starts_with("Sample"))%>%
  separate_rows(`Drug Class`, sep=";", convert = TRUE)%>%
  group_by(`Drug Class`)%>%
  summarise(across(everything(), sum),
            .groups = 'drop')

beta_lactam<-c("cephalosporin", "cephamycin", "penam" , "penem", "carbapenem" ,"monobactam")
penam_penem<-c("penam" , "penem")

ariba_rpkm_beta_lactams<-ariba_rpkm%>%left_join(refname_all_ariba, "ref_name")%>%
  select(`Drug Class`, starts_with("Sample"))%>%
  separate_rows(`Drug Class`, sep=";", convert = TRUE)%>%
  filter(`Drug Class` %in% beta_lactam)%>%
  mutate(`Drug Class` ="beta_lactam")%>%
  unique()%>%
  summarise(across(where(is.numeric), sum))%>%
  mutate(`Drug Class`="beta_lactam")%>%
  select(`Drug Class`, everything())

ariba_rpkm_penam_penem<-ariba_rpkm%>%left_join(refname_all_ariba, "ref_name")%>%
  select(`Drug Class`, starts_with("Sample"))%>%
  separate_rows(`Drug Class`, sep=";", convert = TRUE)%>%
  filter(`Drug Class` %in% penam_penem)%>%
  mutate(`Drug Class` ="penam_penem")%>%
  unique()%>%
  summarise(across(where(is.numeric), sum))%>%
  mutate(`Drug Class`="penam_penem")%>%
  select(`Drug Class`, everything())

##all.equal(colnames(ariba_rpkm_drugclass), colnames(ariba_rpkm_beta_lactams))
all.equal(colnames(ariba_rpkm_drugclass), colnames(ariba_rpkm_penam_penem))
ariba_rpkm_drugclass<-rbind(ariba_rpkm_drugclass, ariba_rpkm_beta_lactams, ariba_rpkm_penam_penem)

```


```{r}
ariba_rar_drugclass<-ariba_rar%>%left_join(refname_all_ariba, "ref_name")%>%
  select(`Drug Class`, starts_with("Sample"))%>%
  separate_rows(`Drug Class`, sep=";", convert = TRUE)%>%
  group_by(`Drug Class`)%>%
  summarise(across(everything(), sum),
            .groups = 'drop')


ariba_rar_beta_lactams<-ariba_rar%>%left_join(refname_all_ariba, "ref_name")%>%
  select(`Drug Class`, starts_with("Sample"))%>%
  separate_rows(`Drug Class`, sep=";", convert = TRUE)%>%
  filter(`Drug Class` %in% beta_lactam)%>%
  mutate(`Drug Class` ="beta_lactam")%>%
  unique()%>%
  summarise(across(where(is.numeric), sum))%>%
  mutate(`Drug Class`="beta_lactam")%>%
  select(`Drug Class`, everything())

ariba_rar_penam_penem<-ariba_rar%>%left_join(refname_all_ariba, "ref_name")%>%
  select(`Drug Class`, starts_with("Sample"))%>%
  separate_rows(`Drug Class`, sep=";", convert = TRUE)%>%
  filter(`Drug Class` %in% penam_penem)%>%
  mutate(`Drug Class` ="penam_penem")%>%
  unique()%>%
  summarise(across(where(is.numeric), sum))%>%
  mutate(`Drug Class`="penam_penem")%>%
  select(`Drug Class`, everything())

##all.equal(colnames(ariba_rar_drugclass), colnames(ariba_rar_beta_lactams))
ariba_rar_drugclass<-rbind(ariba_rar_drugclass, ariba_rar_beta_lactams, ariba_rar_penam_penem)
```

## Generate matrix counts grouped by resistance mechanism
The same problem as antibiotic

```{r}
ariba_rpkm_resmec<-ariba_rpkm%>%left_join(refname_all_ariba, "ref_name")%>%
  select(`Resistance Mechanism`, starts_with("Sample"))%>%
  separate_rows(`Resistance Mechanism`, sep=";", convert = TRUE)%>%
  group_by(`Resistance Mechanism`)%>%
  summarise(across(everything(), sum),
            .groups = 'drop')

ariba_rar_resmec<-ariba_rar%>%left_join(refname_all_ariba, "ref_name")%>%
  select(`Resistance Mechanism`, starts_with("Sample"))%>%
  separate_rows(`Resistance Mechanism`, sep=";", convert = TRUE)%>%
  group_by(`Resistance Mechanism`)%>%
  summarise(across(everything(), sum),
            .groups = 'drop')
```


### Save grouped matrixes

```{r}
save(ariba_rpkm_drugclass, ariba_rpkm_genefamily, ariba_rpkm_resmec, file="output/grouped_rpkm_ariba.RDA")
save(ariba_rar_drugclass, ariba_rar_genefamily, ariba_rar_resmec, file="output/grouped_rar_ariba.RDA")
```


