---
title: "Beta diversity analysis"
author: "Elisa Rubio"
date: "2022-03-06"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(cache=FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE, results ='markup')
options(warn=-1)
```

```{r message=FALSE}
library(tidyverse)
library(vegan)
library(glue)
library(patchwork)
```

## Load summarized and rarefied data

```{r}
load("output/summarized_ariba.RDA")
load("output/summarized_groot.RDA")
load("output/rarefied_ariba.RDA")
load("output/rarefied2500_groot.RDA")
metadata<-read_csv("data/Metadata/metadata.csv")
```


## Create functions

### Bray Curtis Distance from RPKM values or rarefied values (vegdist function)

```{r}
dist_rpkm<-vegdist(t(ariba_rpkm[ ,-1]), method = "bray")
```

### Bray Curtis Distance with avgdist function (performs rarefaction)

```{r}
nCounts_ariba<-apply(ariba_counts[ ,-1], 2, sum)%>%as.data.frame()%>%rename(counts=1)
minCounts<-min(nCounts_ariba$counts)
set.seed(200889)
##dist_rar<- avgdist(t(ariba_counts[ ,-1]), dmethod="bray", sample=minCounts)
```


### Function to calculate all distances

```{r}
dist_func<-function(d_rpkm, d_rar, d_counts, metadata){
  dist_rpkm<-vegdist(t(d_rpkm[ ,-1]), method = "bray")
  dist_rar<-vegdist(t(d_rar[ ,-1]), method = "bray")
  nCounts<-apply(d_counts[ ,-1], 2, sum)%>%as.data.frame()%>%rename(counts=1)
  minCounts<-min(nCounts$counts)
  set.seed(200889)
  dist_avgd<- avgdist(t(d_counts[ ,-1]), dmethod="bray", sample=minCounts)
  
  meta_rpkm<-metadata%>%filter(SampleID %in% colnames(d_rpkm[-1]))
  meta_rar<-metadata%>%filter(SampleID %in% colnames(d_rar[-1]))
  meta_avgd<-metadata%>%filter(SampleID %in% colnames(d_counts[-1]))
  
  d<-list(dist_rpkm, dist_rar, dist_avgd, meta_rpkm, meta_rar, meta_avgd)
  names(d)<-c("dist_rpkm", "dist_rar", "dist_avgd", "meta_rpkm", "meta_rar", "meta_avgd")
  return(d)
}
```


### NMDS
#### NMDS function (for variables with max 3 levels)
```{r}
beta_nmds<-function(dist, metadata, clin_var){
  
  clin_var<-enquo(clin_var)
  clin_var2<-metadata%>%pull(!!clin_var) ##clin_var vector for adonis test
  
  test<-adonis(dist~clin_var2, permutations = 999)
  ptest<-test$aov.tab$`Pr(>F)`[1]
  
  set.seed(200889)
  nmds <- metaMDS(dist)
  
  scores(nmds) %>%
    as_tibble(rownames = "SampleID") %>%
    inner_join(., metadata, by="SampleID") %>%
    ggplot(aes(x=NMDS1, y=NMDS2, color=!!clin_var, fill=!!clin_var)) + 
    stat_ellipse(geom="polygon", show.legend = FALSE, alpha=0.2)+
    geom_point()+
    coord_fixed(ratio = 0.8)+
    labs(caption = glue("ADONIS p value= {ptest}"))+
    scale_color_manual(name=quo_name(clin_var), 
                           values = c("blue", "red","green4"))+ 
        scale_fill_manual(values = c("dodgerblue", "pink", "green"))+
    theme_bw()+
    theme(legend.text = element_text(size=10))}

```


#### NMDS function (for variables with 4-8 levels)
```{r}
beta_nmds2<-function(dist, metadata, clin_var){
  
  clin_var<-enquo(clin_var)
  clin_var2<-metadata%>%pull(!!clin_var) ##clin_var vector for adonis test
  
  test<-adonis(dist~clin_var2, permutations = 999)
  ptest<-test$aov.tab$`Pr(>F)`[1]
  
  set.seed(200889)
  nmds <- metaMDS(dist)
  
  scores(nmds) %>%
    as_tibble(rownames = "SampleID") %>%
    inner_join(., metadata, by="SampleID") %>%
    ggplot(aes(x=NMDS1, y=NMDS2, color=!!clin_var)) + 
    stat_ellipse(show.legend = FALSE)+
    geom_point()+
    coord_fixed(ratio = 0.8)+
    labs(caption = glue("ADONIS p value= {ptest}"))+
    scale_color_manual(name=quo_name(clin_var), 
                           values = c("blue", "red","green4", "orange", "aquamarine4", "magenta2", "gold", "black"))+ 
    theme_bw()+
    theme(legend.text = element_text(size=10))}

```


### PCoA

#### PCoA function (for variables with max 3 levels)

```{r}

beta_pcoa<-function(dist, metadata, clin_var){
  
  clin_var<-enquo(clin_var)
  clin_var2<-metadata%>%pull(!!clin_var) ##clin_var vector for adonis test
  
  test<-adonis(dist~clin_var2, permutations = 999)
  ptest<-test$aov.tab$`Pr(>F)`[1]

  pcoa <- cmdscale(dist, eig=TRUE, add=TRUE)
  positions <- pcoa$points
  colnames(positions) <- c("pcoa1", "pcoa2")
  
  percent_explained <- 100 * pcoa$eig / sum(pcoa$eig)
  
  pretty_pe <- format(round(percent_explained, digits =1), nsmall=1, trim=TRUE)
  
  labels <- c(glue("PCo Axis 1 ({pretty_pe[1]}%)"),
              glue("PCo Axis 2 ({pretty_pe[2]}%)"))
  
  positions %>%
    as_tibble(rownames = "SampleID") %>%
    inner_join(., metadata, by="SampleID") %>%
    ggplot(aes(x=pcoa1, y=pcoa2, color=!!clin_var, fill=!!clin_var)) +
    stat_ellipse(geom="polygon", show.legend = FALSE, alpha=0.2)+
    geom_point() +
    labs(caption = glue("ADONIS p value= {ptest}"), x=labels[1], y=labels[2])+
    coord_fixed(ratio = 0.8)+
    scale_color_manual(name=quo_name(clin_var), 
                           values = c("blue", "red", "gray"))+ 
        scale_fill_manual(values = c("dodgerblue", "pink", "lightgray"))+
    theme_bw()+
    theme(legend.text = element_text(size=10))}
  
```


#### PCoA function (for variables with 4-8 levels)
```{r}
beta_pcoa2<-function(dist, metadata, clin_var){
  
  clin_var<-enquo(clin_var)
  clin_var2<-metadata%>%pull(!!clin_var) ##clin_var vector for adonis test
  
  test<-adonis(dist~clin_var2, permutations = 999)
  ptest<-test$aov.tab$`Pr(>F)`[1]

  pcoa <- cmdscale(dist, eig=TRUE, add=TRUE)
  positions <- pcoa$points
  colnames(positions) <- c("pcoa1", "pcoa2")
  
  percent_explained <- 100 * pcoa$eig / sum(pcoa$eig)
  
  pretty_pe <- format(round(percent_explained, digits =1), nsmall=1, trim=TRUE)
  
  labels <- c(glue("PCo Axis 1 ({pretty_pe[1]}%)"),
              glue("PCo Axis 2 ({pretty_pe[2]}%)"))
  
  positions %>%
    as_tibble(rownames = "SampleID") %>%
    inner_join(., metadata, by="SampleID") %>%
    ggplot(aes(x=pcoa1, y=pcoa2, color=!!clin_var, fill=!!clin_var)) +
    stat_ellipse(show.legend = FALSE)+
    geom_point()+
    coord_fixed(ratio = 0.8)+
    labs(caption = glue("ADONIS p value= {ptest}"))+
    scale_color_manual(name=quo_name(clin_var), 
                           values = c("blue", "red","green4", "orange", "aquamarine4", "magenta2", "gold", "black"))+ 
    theme_bw()+
    theme(legend.text = element_text(size=10))}
```





## Ariba

```{r}
ariba_dist<-dist_func(ariba_rpkm, ariba_rar, ariba_counts, metadata)
```

```{r ariba nmds HIV, message=FALSE, warning=FALSE}
beta_nmds(ariba_dist$dist_rpkm, ariba_dist$meta_rpkm, HIV_Status)+labs(title="Ariba", subtitle = "Bray Curtis distance matrix from RPKM")
beta_nmds(ariba_dist$dist_rar, ariba_dist$meta_rar, HIV_Status)+labs(title="Ariba", subtitle = "Bray Curtis distance matrix from rarefied counts")
beta_nmds(ariba_dist$dist_avgd, ariba_dist$meta_avgd, HIV_Status)+labs(title="Ariba", subtitle = "Bray Curtis distance matrix from rarefied counts (avgdist)")
```

```{r ariba nmds profile}
beta_nmds2(ariba_dist$dist_rpkm, ariba_dist$meta_rpkm, Profile)+labs(title="Ariba", subtitle = "Bray Curtis distance matrix from RPKM")
beta_nmds2(ariba_dist$dist_rar, ariba_dist$meta_rar, Profile)+labs(title="Ariba", subtitle = "Bray Curtis distance matrix from rarefied counts")
beta_nmds2(ariba_dist$dist_avgd, ariba_dist$meta_avgd, Profile)+labs(title="Ariba", subtitle = "Bray Curtis distance matrix from rarefied counts (avgdist)")
```

```{r ariba pcoa hiv}
beta_pcoa(ariba_dist$dist_rpkm, ariba_dist$meta_rpkm, HIV_Status)+labs(title="Ariba", subtitle = "Bray Curtis distance matrix from RPKM")
beta_pcoa(ariba_dist$dist_rar, ariba_dist$meta_rar, HIV_Status)+labs(title="Ariba", subtitle = "Bray Curtis distance matrix from rarefied counts")
```


## Groot

```{r}
groot_dist<-dist_func(groot_rpkm, groot_rar, groot_counts, metadata)
```

```{r groot nmds HIV}
beta_nmds(groot_dist$dist_rpkm, groot_dist$meta_rpkm, HIV_Status)+labs(title="Groot", subtitle = "Bray Curtis distance matrix from RPKM")
beta_nmds(groot_dist$dist_rar, groot_dist$meta_rar, HIV_Status)+labs(title="Groot", subtitle = "Bray Curtis distance matrix from rarefied counts")
beta_nmds(groot_dist$dist_avgd, groot_dist$meta_avgd, HIV_Status)+labs(title="Groot", subtitle = "Bray Curtis distance matrix from rarefied counts (avgdist)")

```

```{r groot pcoa hiv}
beta_pcoa(groot_dist$dist_rpkm, groot_dist$meta_rpkm, HIV_Status)+labs(title="Groot", subtitle = "Bray Curtis distance matrix from RPKM")
beta_pcoa(groot_dist$dist_rar,  groot_dist$meta_rar, HIV_Status)+labs(title="Groot", subtitle = "Bray Curtis distance matrix from rarefied counts")
```

#### Pretty plot

```{r eval=FALSE, include=TRUE}

  set.seed(200889)
  nmds <- metaMDS(ariba_dist$dist_rar)
  
  test<-adonis(ariba_dist$dist_rar~metadata_ariba$HIV_Status, permutations = 999)
  ptest<-test$aov.tab$`Pr(>F)`[1]
  
  scores(nmds) %>%
    as_tibble(rownames = "SampleID") %>%
    inner_join(., metadata, by="SampleID") %>%
    ggplot(aes(x=NMDS1, y=NMDS2, color=HIV_Status, fill=HIV_Status)) + 
    stat_ellipse(geom="polygon", show.legend = FALSE, alpha=0.2)+
    geom_point()+
    coord_fixed(ratio = 0.8)+
    labs(caption = glue("ADONIS p value= {ptest}"))+
    scale_color_manual(name=NULL, 
                       breaks = c("negative", "positive"), 
                       values = c("blue", "red"), 
                       labels=c("HIV negative", "HIV positive"))+
    scale_fill_manual(name=NULL, 
                       breaks = c("negative", "positive"), 
                       values = c("dodgerblue", "pink"), 
                       labels=c("HIV negative", "HIV positive"))+
    theme_bw()+
    theme(legend.text = element_text(size=10))
```
