---
title: "Contingency_table"
author: "Elisa Rubio"
date: "2022-09-03"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(cache=FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE, results ='markup')
options(warn=-1)
```

```{r}
library(tidyverse)
library(gtsummary)
library(kableExtra)
```

```{r}
metadata<-read_csv("data/Metadata/metadata.csv")
```

## Contingecy table: RiskGroup and Microbiome cluster according to HIV status

```{r}
t1<-metadata%>%select(HIV_Status, Cluster, RiskGroup2)%>%
  tbl_summary(by=HIV_Status)%>%
  add_overall()%>%
  add_p()%>%bold_p()%>%
  modify_spanning_header(c("stat_1", "stat_2") ~ "**HIV profile**")

as_kable_extra(t1)%>%kable_paper("striped")
```


```{r}
wide_clust<-metadata%>%count(SampleID, Cluster)%>%
  pivot_wider(names_from = "Cluster", values_from = "n", values_fill = 0)

wide_risk<-metadata%>%count(SampleID, RiskGroup2)%>%
  pivot_wider(names_from = "RiskGroup2", values_from = "n", values_fill = 0)

t2<-metadata%>%
  select(SampleID, HIV_Status)%>%
  inner_join(., wide_clust, by="SampleID")%>%
  inner_join(., wide_risk, by="SampleID")%>%
  select(-SampleID)%>%
  tbl_summary(by=HIV_Status)%>%
  add_overall()%>%
  add_p()%>%bold_p()%>%
  modify_spanning_header(c("stat_1", "stat_2") ~ "**HIV profile**")

as_kable_extra(t2)%>%kable_paper("striped")
```

- HIV positive have more _Bacteroides_ microbiome cluster and HIV negative have more _Prevotella_ microbiome cluster. 

- There are statistcally more msm in the HIV-negative group compared to the HIV-positive group. 

## Contingecy table: RiskGroup and Microbiome cluster

```{r}
t3<-metadata%>%
  select(SampleID, RiskGroup2)%>%
  inner_join(., wide_clust, by="SampleID")%>%
  select(-SampleID)%>%
  tbl_summary(by=RiskGroup2)%>%
  add_overall()%>%
  add_p()%>%bold_p()%>%
  modify_spanning_header(c("stat_1", "stat_2") ~ "**HIV profile**")

as_kable_extra(t3)%>%kable_paper("striped")
```

Most msm have cluster _Prevotella_, and most hts and pwid have cluster _Bacteroides_

