---
title: "Riskgroup2_grouped"
author: "Elisa Rubio"
date: "2022-11-13"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(cache=FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE, results ='markup')
options(warn=-1)
```

```{r message=FALSE}
library(tidyverse)
library(kableExtra)
library(vegan)
library(ggpubr)
library(glue)
library(patchwork)
library(ggstatsplot)
library(broom)
library(ggthemes)
library(ggrepel)
```

#### Load data and functions

```{r}
load("output/summarized_ariba.RDA")
load("output/ariba_dist.RDA")
load("output/grouped_rpkm_ariba.RDA")
load("output/grouped_refname_ariba.RDA")
metadata<-read_csv("data/Metadata/metadata.csv")
source("code/functions_paper.R")
```

## Join HTS with PWID

```{r}
metadata<-metadata%>%
  mutate(MSM_dic=ifelse(RiskGroup2=="msm", "MSM", "no MSM"))
```

### Alpha diversity analysis

```{r}
ariba_alpha_rpkm<-alpha_div(ariba_rpkm, ariba_counts, metadata)
alpha_div_plot_all(ariba_alpha_rpkm, MSM_dic)
```

No differences in alpha diversity 


### Beta diversity analysis

```{r}
ariba_dist_rpkm<-ariba_dist$dist_rpkm
```

Resistome composition is different according to MSM

```{r}
##ECCMID figure
  dist_df<-as.data.frame(as.matrix(ariba_dist_rpkm)); dist_df$SampleID<-rownames(dist_df)
  meta_dist<-inner_join(metadata, dist_df, by="SampleID")%>%
    filter(!is.na(MSM_dic))

  dist<-meta_dist%>%
    select(all_of(.[["SampleID"]]))%>%
    as.dist()

  clin_var2<-meta_dist%>%pull(MSM_dic) ##clin_var vector for adonis test

  test<-adonis(dist~clin_var2, permutations = 999)
  ptest<-test$aov.tab$`Pr(>F)`[1]
  r2test<-round(test$aov.tab$R2[1],2)

  set.seed(200889)
  nmds <- metaMDS(dist, trace = 0, trymax = 200)

  ##scores(nmds) %>%
  nmds$points %>%
    as_tibble(rownames = "SampleID") %>%
    rename(NMDS1=MDS1, NMDS2=MDS2) %>%
    inner_join(., metadata, by="SampleID") %>%
    filter(!is.na(MSM_dic))%>%
    ggplot(aes(x=NMDS1, y=NMDS2, color=MSM_dic, fill=MSM_dic)) +
    stat_ellipse(geom="polygon", show.legend = FALSE, alpha=0.2)+
    geom_point()+
    coord_fixed(ratio = 0.8)+
    annotate("text", x = -0.35, y = 0.22, label = glue("ADONIS p={ptest} \n r2={r2test}"), size=4, hjust=0)+
    scale_color_manual(name="Sexual \nPreference",
                      values = c("red", "gray48"))+
    scale_fill_manual(name="Sexual \nPreference", 
                      values = c("pink", "gray80"))+
    theme_bw()+
    theme(text = element_text(size=15))
```

## Biplot

## Gene family correlations with NMDS axes

```{r}
set.seed(200889)
cor_rpkm_genefamily<-corr_group_envfit(data_group=ariba_rpkm_genefamily, refdata_group = refname_ariba_genefamily, nmds = nmds)

cor_level<-0.2
cor_rpkm_genefamily_corlev2<-cor_rpkm_genefamily%>%filter(p.value<=0.05 & r2 > cor_level)
```


```{r}
cor_rpkm_genefamily_corlev<-cor_rpkm_genefamily_corlev2%>%
   mutate(group_name = replace(group_name, group_name == "23S rRNA with mutation conferring resistance to macrolide antibiotics", "23S rRNA with mutation \n conferring resistance to \n macrolide antibiotics"),
          group_name = replace(group_name, group_name=="Erm 23S ribosomal RNA methyltransferase", "Erm 23S ribosomal \nRNA methyltransferase" ),
          group_name = replace(group_name, group_name=="ABC-F ATP-binding cassette ribosomal protection protein", "ABC-F ATP-binding cassette \nribosomal protection protein"))

biplot_amr_envfit(cor_rpkm_genefamily_corlev, nmds, metadata, MSM_dic, group_name)+
  labs(title = "AMR Gene family")+theme(legend.title = element_blank())
```


## Check differentially abundance AMR

```{r}
amr_sig_rpkm_MSM_dic<-sig_AMR_clin_dic(ariba_rpkm, metadata, refname_all_ariba, MSM_dic)
amr_sig_rpkm_MSM_dic$sig_data%>% kable(caption="** Significant AMR accordig to Sexual preference (Ariba RPKM) **")%>%kable_paper("striped")%>%scroll_box(width = "100%", height = "500px")

write_excel_csv(amr_sig_rpkm_MSM_dic$sig_data, "output/Paper/amr__sexualpreference.xlsx")
```


## Check differentially abundance (grouped by Drug Class)


```{r}
drugclass_sig_rpkm_MSM_dic<-sig_group_clin_dic(ariba_rpkm_drugclass, metadata, refname_ariba_drugclass, MSM_dic)
drugclass_sig_rpkm_MSM_dic$sig_data%>% kable(caption="**Significant Drug Classes, MSM (Ariba RPKM)**")%>%kable_paper("striped")%>%scroll_box(width = "100%", height = "500px")

drugclass<-drugclass_sig_rpkm_MSM_dic$sig_data
write_excel_csv(drugclass, "output/Paper/drugclass_sexualpreference.xlsx")
```


## Check differentially abundance (grouped by Gene Family)

```{r}
genefamily_sexualp<-sig_group_clin_dic(ariba_rpkm_genefamily, metadata, refname_ariba_genefamily, MSM_dic)
genefamily_sexualp$sig_data%>%
  kable()%>%kable_paper("striped")%>%scroll_box(width = "100%", height = "500px")

write_excel_csv(genefamily_sexualp$sig_data, "output/Paper/genefamily__sexualpreference.xlsx")
```











