---
title: "Profile"
author: "Elisa_Linux"
date: "2022-09-10"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(cache=FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE, results ='markup')
options(warn=-1)
```

```{r message=FALSE}
library(tidyverse)
library(kableExtra)
library(vegan)
library(ggpubr)
library(glue)
library(patchwork)
library(ggstatsplot)
library(broom)
```

#### Load data

```{r}
load("output/summarized_ariba.RDA")
load("output/rarefied_ariba.RDA")
load("output/ariba_dist.RDA")
metadata<-read_csv("data/Metadata/metadata.csv")
```

### Load created functions

```{r}
source("code/functions.R")
```


## Joining elite controler + viremic 

```{r}
metadata<-metadata%>%
  mutate(Profile2=ifelse(Profile=="elite_controllers" | Profile=="viremic_controllers", "elite_viremic_controler", Profile ))
```


### Alpha diversity analysis


```{r ariba_alpha_rpkmdiv}
ariba_alpha_rpkm<-alpha_div(ariba_rpkm, ariba_counts, metadata)
alpha_div_plot_all(ariba_alpha_rpkm, Profile2)
```

No differences in alpha diversity according to Profile. 

### Beta diversity analysis

```{r}
ariba_dist_rpkm<-ariba_dist$dist_rpkm
beta_nmds2(ariba_dist_rpkm, metadata, Profile2)
```

Adonis test significant but not clear cluster (pairwise test?)


## Divide according to CD4 values:

```{r}
gghistostats(
  data = metadata,
  x = CD4, 
  xlab = "CD4 levels",
  title = "Distribution of CD4",
  binwidth = 50,
  bin.args = list(color = "black", fill = "green4", alpha = 0.7))+
  geom_vline(xintercept = 400, colour="red", linetype="dashed")
```


```{r}
CD4_cutoff=400

metadata<-metadata%>%
  mutate(CD4_hl=ifelse(CD4>=CD4_cutoff, "high", "low"),
         CD4_hl=ifelse(HIV_Status=="negative", "high", CD4_hl),
         CD4_hl=factor(CD4_hl, levels = c("low", "high")))
```

### Diversity analysis


```{r ariba_alpha_rpkmdiv}
ariba_alpha_rpkm<-alpha_div(ariba_rpkm, ariba_counts, metadata)
p1<-alpha_div_plot_all(ariba_alpha_rpkm, CD4_hl)
p2<-beta_nmds(ariba_dist_rpkm, metadata, CD4_hl)
p1/p2
```

No differences in alpha and beta diversity according to CD4 levels.

## Analyze only msm group

```{r}
metadata_msm<-metadata%>%
  filter(RiskGroup2=="msm")
```

```{r}
gghistostats(
  data = metadata_msm,
  x = CD4, 
  xlab = "CD4 levels",
  title = "Distribution of CD4",
  binwidth = 50,
  bin.args = list(color = "black", fill = "green4", alpha = 0.7))+
  geom_vline(xintercept = 400, colour="red", linetype="dashed")
```

```{r}
metadata_msm<-metadata_msm%>%
  mutate(CD4_hl=ifelse(CD4>=CD4_cutoff, "high", "low"),
         CD4_hl=ifelse(HIV_Status=="negative", "high", CD4_hl),
         CD4_hl=factor(CD4_hl, levels = c("low", "high")))
```

### Diversity analysis

```{r}
ariba_alpha_rpkm<-alpha_div(ariba_rpkm, ariba_counts, metadata_msm)
p1<-alpha_div_plot_all(ariba_alpha_rpkm, CD4_hl)
p2<-beta_nmds(ariba_dist_rpkm, metadata_msm, CD4_hl)
(p1/p2)+plot_annotation(title = "MSM subsample")
```

No differences in alpha and beta diversity according to CD4 levels in msm patients

## Test each AMR gene separatedly for a dictomic clinical variable:

```{r}
data<-ariba_rpkm
metadata
refdata<-refname_all_ariba

sig_AMR_clin_dic<-function(data, metadata, refdata, clin_var){
clin_var<-enquo(clin_var) 

##transpose data
  ref_name<-pull(data, ref_name)
  data<-as_tibble(cbind(SampleID = names(data), t(data)))%>%slice(-1)%>%
    mutate_at(vars(-("SampleID")),as.numeric)
  colnames(data)<-c("SampleID", ref_name)

##merge data with metadata and reference data    
  data_all<-data%>%
    pivot_longer(-SampleID, names_to = "ref_name", values_to = "value")%>%
    inner_join(., metadata, by="SampleID")%>%
    inner_join(., refdata, by="ref_name")
  
  sig_amr<-data_all%>%
    nest(data=-ref_name)%>%
    mutate(test=map(.x=data, ~wilcox.test(value~!!clin_var, data=.x)%>%tidy))%>%
    unnest(test)%>%
    mutate(p.adjust=p.adjust(p.value, method = "BH"))%>%
    filter(p.adjust<0.05)
  return(sig_amr)}
  
```

```{r}
sig_AMR_clin_dic(ariba_rpkm, metadata, refname_all_ariba, CD4_hl)
```

